---
title: "Do More Subdivisions in Wards Increase Voter Turnout? An Analysis of the 2022 Toronto Municipal Election"
author: "Inessa De Angelis"
thanks: "Code and data from this analysis are available at: https://github.com/InessaDeAngelis/Toronto_Elections" 
date: "today"
date-format: "long" 
abstract: "The 2022 Toronto Municipal Election witnessed the lowest voter turnout in the city’s history since amalgamation in 1997, with voter turnout on a ward-by-ward basis ranging from 22% to 38%. Using data from OpenDataToronto, this paper disproves the hypothesis that having more subdivisions within a ward bolsters voter turnout. This paper reveals that more subdivisions within a ward does not necessarily correlate to a higher voter turnout."
format: pdf
toc: true
number-sections: true
bibliography: ref.bib
---

```{r}
#| message: false
#| echo: false

#### Workspace set-up ####

## Load packages
library(tidyverse)
library(knitr)
library(kableExtra)
library(here)
library(ggplot2)
library(ggbeeswarm)
library(ggrepel)

#read in cleaned data #
cleaned_voter_statistics <- read.csv(here::here("inputs/data/cleaned_voter_statistics.csv"))
show_col_types = FALSE

#read in cleaned data #
summarized_voter_statistics <- read.csv(here::here("inputs/data/summarized_voter_statistics.csv"))
show_col_types = FALSE

#read in cleaned data #
summarized_voter_turnout <- read.csv(here::here("inputs/data/summarized_voter_turnout.csv"))
show_col_types = FALSE

#read in cleaned data #
finalized_voter_statistics <- read.csv(here::here("inputs/data/finalized_voter_statistics.csv"))
show_col_types = FALSE
```

## Introduction
As mandated by the Government of Ontario, residents of the City of Toronto went to the polls on October 24, 2022 to elect a mayor, councillors, and school board trustees. Then-sitting Mayor, John Tory, sought re-election, along with a number of incumbent city councillors. Coming out of the height of the Covid-19 pandemic, the 2022 election largely upheld the status quo and did not feature ambitious policy platforms. Voter turnout across the city fell to 30% - the lowest in the city’s history since amalgamation in 1997, with turnout ranging on a ward-by-ward basis from 22% to 38% [@Turnout22; @DismalTurnout]. 

Individual turnout by ward depended on a number of factors, including the accessibility of polling locations within subdivisions to cast a ballot. The City of Toronto @ElectionDictionary defines a ward as “a geographical area represented by a member of Council.” Following Premier Doug Ford’s decision in the summer of 2018 to cut down the number of wards in the city to match federal and provincial riding boundaries, there are currently 25 wards in the City of Toronto [@BigCityElxns]. Subdivisions are defined by the City of Toronto [@VoterStatistics] as “.... geographical area[s within a ward] designated by the City Clerk.” Previous studies concentrating on large cities in the United States, including Atlanta revealed that having polling locations in close proximity to a voter’s home bolsters turnout and even minor changes in placement of a polling location can have significant impact on a voter’s decision to cast a ballot [@Locationx3]. There has been little research into the impact of polling location placement in relation to election turnout, especially in Canadian cities like Toronto and this paper will contribute to the investigation of this phenomenon. 

To examine the impact of subdivision placement on voter turnout, this paper is organized into the following sections: Data, Results, Discussion, and Conclusion.  In the Data section, I discuss the nature of the spreadsheet obtained through the City of Toronto’s OpenDataToronto Library [@rOpenDataToronto] and the steps I took to clean and analyze the data. The Results section highlights trends found during the analysis process, while the Discussion section further evaluates the trends and presents insight. Lastly, the Conclusion section summarizes the main findings from this paper. 

## Data
The data utilized throughout this paper was obtained through the City of Toronto’s OpenDataToronto Library [@rOpenDataToronto]. The data set used is entitled 'Elections - Voter Statistics' [@VoterStatistics]. Data was collected and analyzed using the statistical programming software R [@r], with additional support from `tidyverse` [@rTidyverse], `ggplot2` [@rGgplot2], `dplyr` [@rDplyr], `readr` [@rReadr], `tibble` [@rTibble], `janitor` [@rJanitor], `KableExtra` [@rKableExtra], `knitr` [@rknitr], `ggbeeswarm` [@rggbeeswarm], `ggrepel` [@rggrepel], and `here` [@rHere]. A further discussion of the data collection, cleaning, and analysis process can be found later on in this paper. 

### Election Voter Statistics
This dataset, published by the City Clerk's Office [@VoterStatistics] outlines voter statistics on a ward-by-ward and entire city basis for the 2022 municipal election. For each subdivision within a ward, the data set shows the polling location name and address, number of additions and corrections to the voter's list, number of eligible electors and number who voted, and rejected and declined ballots. This data set was last refreshed on February 7, 2023 and captured for this paper on May 21, 2023. 

Upon analysis, columns such as rejected and declined ballots were deemed beyond the scope of this paper. I conducted the first step of basic data cleaning to eliminate additional columns and simplify the names of retained columns (see @tbl-cleaned_voter_statistics).

```{r}
#| message: false
#| echo: false
#| label: tbl-cleaned_voter_statistics
#| tbl-cap: Sample of Cleaned Voter Statistics Data

head (cleaned_voter_statistics, 5) |>
  kable(
    col.names = c("Ward", "Sub", "Total Eligible Electors",	"Number Voted",	"Percent Voted"),
  booktabs = TRUE
  )
```
### Number of Subdivisions 

The Election Voter Statistics data set [@VoterStatistics] listed "NA" for the number of subdivisions per ward, but included the voter turnout average for all subdivisions within a ward. Consequently, I wrote a script using the `count` function to count the number of subdivisions per ward. I included the Advance Vote and Mail In Voting subdivisions when counting the total number of subdivisions per ward (see @tbl-summarized_voter_statistics). Firstly, there are only two Advance Vote locations distributed across an entire ward, meaning residents have to be motivated to travel farther away from their home to cast their ballot. Secondly, Mail In Voting requires applying online or by phone through Toronto Elections and then returning the completed Mail In Voting package by dropping it in a Canada Post mailbox or hand delivering it to one of the select Toronto Elections drop boxes [@ElxnAccessibility]. Despite the number of subdivisions per ward, the City of Toronto has earmarked subdivision 97 for Mail In Voting, while subdivisions 98 and 99 are designated Advance Vote locations.

```{r}
#| message: false
#| echo: false
#| label: tbl-summarized_voter_statistics
#| tbl-cap: Sample of Summarized Voter Statistics Data

head(summarized_voter_statistics, 5) |>
  kable(
    col.names = c("Ward", "Number of Subdivisions"), 
    booktabs = TRUE
  )
```

### Turnout by Ward
Moreover, using the cleaned voter statistics data set, I added the voter turnout by ward percentage calculated by the Toronto Elections from the bottom of each ward section to a table I created with each ward number. To view a sample of this data set, please see @tbl-summarized_voter_turnout. 


```{r}
#| message: false
#| echo: false
#| label: tbl-summarized_voter_turnout
#| tbl-cap: Sample of Summarized Voter Turnout Data

head(summarized_voter_turnout, 5) |>
  kable(
    col.names = c("Ward", "Turnout (%)"), 
    booktabs = TRUE
  )
```

### Merged Number of Subdivisions and Turnout 
Lastly, using the `merge` function, I created a script to combine the summarized number of subdivisions and voter turnout data into one data set. See @tbl-finalized_voter_turnout for a sample of this merged data set. 

```{r}
#| message: false
#| echo: false
#| label: tbl-finalized_voter_turnout
#| tbl-cap: Sample of Merged Subdivisions and Voter Turnout Data Set

head(finalized_voter_statistics, 5) |>
  kable(
    col.names = c("Ward", "Turnout (%)", "Subdivisions"), 
    booktabs = TRUE
  )
```

## Results
### Subdivisions per Ward
During the 2022 municipal election, there were 1,535 subdivisions across Toronto's 25 wards. Ward 10 (Spadina-Fort York) had the highest number of subdivisions at 94, while Ward 23 (Scarborough North) had the lowest number at 38. Ward 13 (Toronto Centre) had the second highest number of subdivisions at 90, with Ward 3 (Etobicoke-Lakeshore) coming in third highest with 83 subdivisions. Ward 7 (Humber River-Black Creek) had the second lowest number of subdivisions at 47, following by Ward 25 (Scarborough-Rouge Park) with 48. @fig-voter_statistics displays the total number of subdivisions for all 25 wards. 

```{r}
#| message: false
#| echo: false
#| label: fig-voter_statistics
#| fig-cap: Number of Subdivisions per ward

#Code refernced from: https://stackoverflow.com/questions/11335836/increase-number-of-axis-ticks
finalized_voter_statistics |>
 ggplot(mapping = aes(x = ward, y = n)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 25)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 25))+
  labs(
    title = "Number of Subdivisions per Ward",
    x = "Ward",
    y = "Number of Subdivisions",
  ) +
  theme_classic() 
```

### Voter Turnout per Ward
Voter turnout across the entire city during the 2022 municipal election fell to 30%, with Ward 4 (Parkdale-High Park) having the highest turnout at 38% and Ward 7 (Humber River-Black Creek) having the lowest at 22%. Ward 19 (Beaches-East York) and Ward 14 (Toronto Danforth) were tied for second highest turnout at 36% each. The third highest turnout at 34% was in Ward 12 (Toronto-St. Paul's). The wards with the second lowest turnout were Ward 1 (Etobicoke North) and Ward 10 (Spadina-Fork York) at 24% each. Ward 6 (York Centre) and Ward 23 (Scarborough North) with 25% turnout respectively were tied for third lowest turnout. For a full picture of voter turnout per ward, please refer to @fig-turnout_by_ward. 

```{r}
#| message: false
#| echo: false
#| label: fig-turnout_by_ward
#| fig-cap: Voter Turnout by ward

#Voter turnout by ward
#Code refernced from: https://stackoverflow.com/questions/11335836/increase-number-of-axis-ticks
finalized_voter_statistics |>
 ggplot(mapping = aes(x = ward, y = turnout)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 25)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 25))+
  geom_bar(stat = "identity") +
  labs(
    title = "Voter Turnout by Ward",
    x = "Ward",
    y = "Turnout (%)",
  ) +
  theme_classic() 
```

```{r}
#| message: false
#| echo: false
#| label: fig-comparing-sub-turnout
#| fig-cap: Correlation between Number of Subdivisions and Voter Turnout

#Voter turnout by ward
#Code referenced from: https://github.com/christina-wei/INF3104-1-Covid-Clinics/blob/main/outputs/paper/covid_clinics.pdf
# & https://stackoverflow.com/questions/66652330/error-ggplot-continuous-x-aesthetic-did-you-forget-aesgroup
finalized_voter_statistics |>
  ggplot(aes(x = turnout, y = n, label = ward)) +
  # ggplot(aes(x = factor(turnout), y = n, label = ward)) +
  geom_point() + 
  # geom_point(aes(group = turnout)) + 
  # geom_beeswarm() +
  # geom_smooth(method="lm")+
  # geom_text(hjust=-05, vjust=0.5) +
  ggrepel::geom_text_repel() + 
labs(
    title = "Number of Subdivisions vs Voter Turnout by Ward",
    x = "Turnout (%)",
    y = "Number of subdivisions",
    label = "Ward"
  ) +
  theme_classic()
```

## Discussion
Overall, the data reveals that having more subdivisions within wards does not automatically correlate to higher voter turnout, as illustrated by @fig-comparing-sub-turnout. Ward 4 (Parkdale-High Park) which had the highest voter turnout at 38% has 70 subdivisions within the ward, which falls approximately in the middle of the number of subdivisions across all 25 wards. Ward 19 (Beaches-East York) and Ward 14 (Toronto Danforth) which both saw the second highest voter turnout at 36% had 57 and 52 subdivisions respectively, once again highlighting that more subdivisions does not necessarily correlate to higher voter turnout. 

Furthermore, Ward 7 (Humber River-Black Creek) had a 22% voter turnout and 47 subdivisions, emphasizing that having fewer subdivisions potentially contributed to having the lowest voter turnout across the city. Ward 10 (Spadina-Fork York), which had the second lowest voter turnout at 24%, accounts for the highest number of subdivisions within a ward at 94. Previous research reveals that younger, more transient communities reside in the Spadina-Fork York riding and are less likely to engage with municipal politics, potentially contributing to the low voter turnout despite the high number of subdivisions [@Turnout22].

The wards within Scarborough (Wards 20-25) consistently had some of the lowest voter turnouts and fewest number of subdivisions across all 25 wards. Ward 23 (Scarborough North) witnessed the third lowest voter turnout across the entire city at 25% and the fewest number of subdivisions at 38. Ward 20 (Scarborough Southwest) had the highest turnout of all the Scarborough wards at 32% and had 61 subdivisions. Scarborough's recognized challenges with accessible public transit, paired with fewer subdivisions and polling locations potentially contribute to the lower voter turnout [@transitdesert]. 

Within the data set used for analysis, there are several potential measurement errors. Firstly, the raw data set listed the total number of subdivisions per ward as "NA", despite other aspects of voter turnout and subdivisions being calculated by Toronto Elections. Secondly, there were gaps in the number and corresponding numbering of subdivisions which were not explained by the City Clerk or Toronto Elections. For example, Ward 1 is listed as having 60 subdivisions, plus Advanced Vote and Mail In Voting locations, but subdivisions 4, 20, 43-46, 50, and 58 are unaccounted for, totaling 55 subdivisions. Thirdly, there is a lack of clarity whether the people who voted at Advanced Polls or by Mail In Ballot were removed from the total number of electors by subdivision, since these individuals did not vote at their assigned subdivision. If voters cast their ballot early and were not removed from the total number of electors at their specific subdivision, this could impact how voter turnout by subdivision is calculated. 

There are several limitations with conducting this analysis. The creation of subdivision boundaries within wards lacks transparency and clarity, while impacting the local democratic process. The City Clerk is responsible for drawing up all subdivisions, however the data and information they use to make these decisions is not public and it is challenging to account for any bias introduced during this process. Toronto Elections [@ElxnAccessibility] clearly outlines their process for finding and establishing accessible polling locations within subdivisions but does not elaborate on the process of establishing subdivisions themselves. 

Moreover, the number of subdivisions and polling location placement alone cannot dictate voter behaviour, as there are many other socio-economic factors which inform why an elector does or does not cast a ballot. The caliber of local candidates and the nature of the election, whether an incumbent is standing for re-election or the seat is open for a new councillor impacts voter turnout and should be accounted for in future analysis. 


## Conclusion
This paper investigates the number of subdivisions within Toronto’s 25 wards in relation to voter turnout to see if the wards with more subdivisions had higher turnout during the 2022 municipal election. My results reveal that wards with more subdivisions did not correlate to having higher voter turnout. Factors such as income, level of education attained, and an analysis of local races for councillor and school board trustee within a ward should be analyzed in further studies of voter behaviour during municipal elections. Data from the upcoming Toronto Mayoral by-election on June 26, 2023 may reveal further insights into the placement of polling locations in relation to voter turnout. Cities and municipalities across Ontario are facing declining voter turnout and the factors must be identified and addressed, whether it be more strategically choosing subdivisions and polling locations in wards, introducing online voting mechanisms, or implementing democratic reform [@reducingcost]. Strong local representation is crucial for the future of democracy in Toronto and subdivisions play an important role in ensuring voters have their voices heard. 

## References
